---
title: "RMDL Assessment 2: Data Analysis Report"
authors:
  - name: B288421
bibliography: references.bib
execute: 
  echo: true
---

## Background

The Stroop task, first described by @stroop_studies_1935, is one of the most well known phenomenons in cognitive psychology. In a classic Stroop task setup, participants are presented with colored words printed in colored ink and asked to name the ink color while ignoring the word itself. The key finding is that participants are slower when the word and ink color are incongruent compared to when they are congruent [@macleod_stroop_2006]. This difference in reaction time (RT) is known as the Stroop interference effect and is thought to reflect the automaticity of word reading [@macleod_half_1991]. In other words, because skilled readers process written words automatically, the meaning of the colored word interferes with the identification of its ink color.

The Stroop effect has also been studied in bilingual speakers. @macleod_half_1991 reviews approximately two dozen studies on this topic. He notes that a dominant language has greater potential for interference than a nondominant one, and that the pattern of interference shifts as proficiency in the L2 increases. This suggests that the degree of Stroop interference in a given language reflects the extent to which word reading has become automatized in that language. Therefore, if word reading is less automatic in L2, the color word should interfere less with ink color identification, resulting in a smaller congruency effect.

### Research Questions/Hypotheses

**Research questions**

-   **RQ 1:** Are participants slower when responding to incongruent trials compared to congruent trials?
-   **RQ 2:** Does the Stroop interference effect differ between L1 and L2 English speakers?

**Hypotheses**

-   **RH 1:** Participants will show slower reaction times on incongruent trials compared to congruent trials, reflecting the Stroop interference.
-   **RH 2:** Since automatic word reading is stronger in a first language, L1 English speakers will show a larger Stroop interference effect compared to L2 English speakers.

## Methods

### Participants

25 participants took part in the study, and they were divided into L1 English speakers (n = 11) and L2 English speakers (n = 14) based on their self report. Participants were recruited through convenience sampling. No formal exclusion criteria were applied.

### Materials

The experiment was designed and administered using Testable. The stimuli consisted of three color words (RED, BLUE, and GREEN) presented in colored ink (red, blue, or green). Each word appeared in either a congruent condition (word and ink color match) or an incongruent condition (word and ink color do not match). Each block contained 20 critical trials (10 congruent, 10 incongruent) and 12 filler trials consisting of non-color words (CAR, SHOE, HOUSE) presented in colored ink.

### Procedure

Participants completed the experiment online via Testable. They were instructed to respond to the ink color of the word by pressing the **LEFT** arrow key for red, the **UP** arrow key for blue, and the **RIGHT** arrow key for green as quickly and accurately as possible. Following the experimental trials, participants completed a short questionnaire asking whether English was their first language, their self-rated English proficiency, and whether they currently reside in an English-speaking country.

### Statistical analysis

Data were analyzed using Bayesian multilevel modeling in R using the `brms` package. Only critical trials were retained, and RT outliers were removed (RT \< 200ms or \> 2000ms). A model was fitted with log RT as the outcome, with congruency, English language status, and their interaction as fixed effects, and varying intercepts and slopes for congruency by participant. The main effect of congruency was used to assess whether a Stroop effect was present, and the interaction term was used to assess whether the Stroop effect differed between L1 and L2 speakers.

### Open research statement

The experimental trial file, raw data, and analysis code are available at \[<https://github.com/ww2526852/rdml-assignment2-repo>\]. The experiment was administered via Testable and can be accessed at \[<https://tstbl.co/661-427>\].

## Data

### Read and process the data

The data was downloaded from Testable in long format, with all 25 participants' trial results combined into a single CSV file. The file was read into R using read_csv().

Several processing steps were applied to prepare the data for analysis. First, each participant's response to the question "Is English your first language?" (row 82 in the trial file) was extracted and used to create a new column **english_status**, coded as **L1** for participant who answered "Yes" and **L2** for all other responses. This column identified the between-subjects grouping variable for the analysis.

The data was then filtered to retain only critical test trials (**condition1 == "critical"**), excluding practice and filler trials that are not relevant to the research question. RT outliers were removed, excluding trials with RT below 200ms (anticipatory responses) or above 2000ms (inattentive responses).

The relevant columns in the cleaned dataset are as follows:

-   **filename:** unique identifier for each participant, based on their Testable result file
-   **condition2:** the congruency condition of each trial
-   **RT:** reaction time in milliseconds, measured from stimulus onset to key press
-   **correct:** accuracy of the responses, coded as 1 (correct) or 0 (incorrect)
-   **english_status:** English language status of the participant, coded as L1 (English is their first language) or L2 (English is not their first language)

```{r}
#| label: read-data

library(tidyverse)

#read raw data file
raw <- read_csv("data/661427_260428_102656_long.csv")

#extract rowNo 82 (L1 question) for each participant
l1_info <- raw |>
  filter(rowNo == 82) |>
  select(filename, l1_response = response)

#merge l1_info and create english_status column
raw <- raw |>
  left_join(l1_info, by = 'filename') |>
  mutate(
    english_status = ifelse(l1_response == 'Yes', 'L1', 'L2')
  )

#filter for critical trials and select relevant columns 
cleaned_data <- raw |>
  filter(type == 'test', condition1 == 'critical') |>
  select(filename, stim1, textColor, condition2, RT, correct, english_status) |>
  mutate(
    condition2 = factor(condition2, levels = c('congruent', 'incongruent')),
    english_status = factor(english_status, levels = c('L1', 'L2'))
  ) |>
  filter(RT > 200, RT < 2000)

```

{{< pagebreak >}}

### Plot the data

#### Plot 1: RT distribution

*Figure 1* is a density plot showing the distribution of log-transformed reaction times across congruency conditions and is faceted by *english_status.* The x-axis shows log RT and the y-axis shows density. Congruency condition is indicated by fill color, with congruent trials shown in pink and incongruent trials shown in teal. The left panel shows L1 English speakers and the right panel shows L2 English speakers.

```{r}
library(ggplot2)

cleaned_data |>
  ggplot(aes(x=log(RT),fill=condition2)) + 
  geom_density(alpha=0.5) +
  facet_wrap(~english_status) +
  labs(
    x = 'log RT',
    caption = 'Figure 1: Distribution of log RT by congruency and English status',
    fill = 'Congruency'
  )
```

In both panels, the incongruent distribution is shifted rightward relative to the congruent distribution, indicating that participants were generally slower on incongruent trials regardless of language background, and this is consistent with a Stroop interference effect. The shift appears visually larger in the L1 panel compared to the L2 panel, suggesting that the interference effect may be greater for L1 English speakers.

#### Plot 2: Mean log RT

*Figure 2* is a point plot showing mean log-transformed reaction time for each congruency condition, with congruency on the x-axis and mean log RT on the y-axis. English status is indicated by color, with L1 speakers shown in pink and L2 speakers in teal.

```{r}

cleaned_data |>
  group_by(english_status,condition2) |>
  summarize(
    mean_logRT = mean(log(RT)),
    sd_logRT = sd(log(RT))
  ) |>
  ggplot(aes(x = condition2, y = mean_logRT,color = english_status, group = english_status)) +
  geom_point(size = 3) +
  geom_line()+
  geom_errorbar(aes(ymin = mean_logRT - sd_logRT,
                    ymax = mean_logRT + sd_logRT),
                width = 0.1) +
  labs(
    caption = 'Figrue 2: Mean log RT by congruency and English status',
    x = 'Congruency',
    y = 'Mean log RT',
    color = 'English status'
  )

```

Both groups show a clear increase in mean log RT from congruent to incongruent trials, consistent with a Stroop interference effect in both L1 and L2 English speakers. The slope of the increase appears steeper for L1 speakers than L2 speakers, suggesting a larger congruency effect for the L1 group, which is consistent with the predicted direction. L2 speakers show higher overall log RT in both conditions, suggesting they are generally slower to respond.

#### Plot 3: Accuracy

*Figure 3* is a bar plot showing the mean accuracy for each combination of congruency condition and English status. Congruency is shown on the x-axis and mean accuracy on the y-axis, ranging from 0 to 1. English status is indicated by fill color, with L1 speakers in pink and L2 speakers in teal. Bars for each group are displayed side by side within each congruency condition.

```{r}

cleaned_data |>
  group_by(english_status,condition2) |>
  summarize(mean_acc = mean(correct)) |>
  ggplot(aes(x = condition2,y = mean_acc,fill = english_status)) +
  geom_col(position = 'dodge') +
  labs(
    caption = 'Figure 3: Accuracy by congruency and English status',
    x = 'Congruency',
    y = 'Mean accuracy',
    fill = 'English status'
  ) +
  ylim(0,1)
```

Accuracy is uniformly high across all conditions, with all bars approaching 1.0. A small decrease in accuracy is visible from congruent to incongruent trials in both groups, consistent with the expected direction of a Stroop interference effect. The difference between L1 and L2 speakers within each congruency condition is minimal.

{{< pagebreak >}}

### Summarize the data

The table below reports mean RT, SD of RT, mean log RT, SD of log RT, mean accuracy, and number of trials for each combination of congruency condition and English status groups.

```{r}
summary_stats <- cleaned_data |>
  group_by(english_status,condition2) |>
  summarize(
    mean_RT = mean(RT),
    sd_RT = sd(RT),
    mean_logRT = mean(log(RT)),
    sd_logRT = sd(log(RT)),
    accuracy = mean(correct),
    n_trials = n()
  )

summary_stats
```

L1 English speakers showed a mean RT of 607ms on congruent trials and 742ms on incongruent trials, reflecting a Stroop interference effect of approximately 134ms. L2 English speakers showed a mean RT of 677ms on congruent trials and 770ms on incongruent trials, reflecting a small interference effect of approximately 94ms. Both groups show the expected pattern of slower response on incongruent trials. The Stroop effect appears numerically larger for L1 speakers than L2 speakers, consistent with the predicted direction. Accuracy was high across all conditions, exceeding 0.94 in all groups, indicating a ceiling effect.

{{< pagebreak >}}

## Regression modelling

To investigate whether English status modulates the Stroop interference effect, a Bayesian multilevel model was fitted using the `brms` package in R, using a Gaussian family. Log-transformed RT was used as the outcome variable. The model included congruency (congruent vs. incongruent), English status (L1 vs. L2), and their interaction as fixed effects.

The interaction term is of primary interest, as it tests whether the size of the Stroop effect differs between L1 and L2 English speakers. Varying intercepts and varying slopes for congruency were included by participant, allowing baseline RT and the congruency effect to differ across individuals.

```{r}
library(brms)

rt_bm <- brm(
  log(RT) ~ condition2 * english_status + 
    (1 + condition2 | filename),
  family = gaussian,
  data = cleaned_data,
  cores = 4,
  seed = 1234,
  file = "data/cache/rt_bm"
)

summary(rt_bm)

```

{{< pagebreak >}}

The intercept represents the estimated baseline log RT for L1 English speakers on congruent trials (6.35, 95% CrI\[6.25, 6.45\]). The effect of congruency for L1 speakers was 0.18 log RT units (95% CrI \[0.12, 0.25\]), indicating that incongruent trials were credibly slower than congruent trials, as the 95% CrI excludes 0. L2 speakers were estimated to be slightly slower overall than L1 speakers (0.11, 95% CrI\[-0.03, 0.24\]), though this difference was not conclusive as the CrI includes 0.

The interaction term between congruency and English status (-0.06, 95% CrI\[-0.15, 0,.03\]) is of primary interest because it tests whether the Stroop effect differs between groups. The negative estimate indicates that L2 speakers show a smaller Stroop interference effect than L1 speakers, consistent with the predicted direction. However, the 95% CrI includes 0, meaning that there is no conclusive evidence that the two groups differ in the magnitude of the Stroop effect. The random effect indicates meaningful variability in baseline RT across participants (sd(Intercept) = 0.15), confirming that individuals differ in their overall response speed, while variability in the Stroop effect across participants was small (sd(condition2incongruent) = 0.05).

{{< pagebreak >}}

### Model plots

*Figure 4* is a point plot showing the model-estimated mean log RT for congruent and incongruent trials, with congruency on the x-axis and log RT on the y-axis. English status is indicated by color, with L1 speakers in pink and L2 speakers in teal.

```{r}

plot(
  conditional_effects(rt_bm, 'condition2:english_status'),
  points = FALSE
) 
```

*Figure 4: Model-estimated log RT by congruency and English status*

Both groups show an increase in model-estimated log RT from congruent to incongruent trials, consistent with a Stroop interference effect in both L1 and L2 English speakers. The increase appears slightly larger for L1 speakers than L2 speakers, reflecting the numerically larger interference effect observed in the model coefficients. The credible intervals overlap substantially between groups, particularly on incongruent trials, which is consistent with the interaction term whose 95% CrI included 0, indicating uncertainty about whether the groups genuinely differ in the magnitude of the Stroop effect.

*Figure 5* is a density plot showing the posterior distributions of the Stroop effect for L1 speakers (green), L2 speakers (blue), and the difference between the two groups (pink). The x-axis show the estimated Stroop effect in log RT units and the y-axis shows density. The vertical black line at 0 serves as a reference point: distributions entirely to the right of this line indicates a credibly positive effect, while distributions crossing the line indicate uncertainty about the direction of the effect.

```{r}
rt_draws <- as_draws_df(rt_bm, '^b', regex = TRUE)

rt_draws <- rt_draws |>
  mutate(
    stroop_L1 = b_condition2incongruent,
    stroop_L2 = b_condition2incongruent + `b_condition2incongruent:english_statusL2`,
    group_difference = stroop_L2 - stroop_L1
  )

rt_stroop_diff <- rt_draws |>
  select(stroop_L1, stroop_L2, group_difference) |>
  pivot_longer(everything())

rt_stroop_diff |>
  ggplot(aes(value, fill = name)) +
  geom_vline(xintercept = 0) +
  geom_density(alpha = 0.5)
```

*Figure 5: Posterior distributions of the Stroop effect for L1 and L2 speakers and their difference*

The posterior distributions for both **stroop_L1** and **stroop_L2** lie entirely to the right of 0, confirming that both groups show a credible Stroop interference effect. The **stroop_L1** distribution is shifted slightly further right than the **stroop_L2** distribution, and the **group_difference** distribution is centered to the left of 0, indicating that the Stroop effect is estimated to be smaller for L2 than L1 speakers, which is consistent with the predicted direction. However, the distribution crosses 0, reflecting uncertainty about whether this difference is genuine.

```{r}

rt_stroop_diff |>
  group_by(name) |>
  summarise(
    q5  = round(quantile(value, 0.05), 3),
    q10 = round(quantile(value, 0.10), 3),
    q20 = round(quantile(value, 0.20), 3),
    q80 = round(quantile(value, 0.80), 3),
    q90 = round(quantile(value, 0.90), 3),
    q95 = round(quantile(value, 0.95), 3)
  )
```

The Stroop effect for L1 English speakers was credibly positive, with a 90% CrI of \[0.142, 0.226\] and a 95% CrI of \[0.128, 0.238\], both excluding 0. This provides clear evidence of a Stroop interference effect in L1 English speakers. Similarly, the Stroop effect for L2 English speakers was also credibly positive, with a 90% CrI of \[0.084, 0.160\] and 95% CrI of \[0.074, 0.172\], both excluding 0, indicating a Stroop interference effect in L2 English speakers as well.

For the group difference, the 90% of CrI is \[-0.119, -0.004\], which excludes 0, suggesting that at 90% probability the Stroop effect is smaller for L2 speakers than L2 speakers. However, the 95% CrI \[-0.136, 0.013\] includes 0, indicating uncertainty about whether the groups genuinely differ at 95% confidence. The results are therefore inconclusive as to whether English language status modulates the Stroop interference, though the direction of the difference is consistent with the prediction than L1 English speakers show greater interference than L2 English speakers.

## Discussion

The results provided clear evidence in support of RH 1. Both L1 and L2 English speakers showed credibly slower reaction times on incongruent trials compared to congruent trials. This is consistent with the well-established Stroop interference effect, whereby the automatic reading of a colored word interferes with the identification of its ink color.

With respect RQ 2, the results were inconclusive. The group difference was in the predicted direction, with L1 English speakers showing a numerically larger Stroop effect than L2 English speakers, and the 90% CrI for the difference excluded 0 \[-0.119, -0.004\], providing weak evidence consistent with RH 2. However, the 95% CrI included 0 \[-0.136, 0.013\], meaning the data are also compatible with no group difference. This inconclusiveness can likely be attributed in part to the small sample size, which limits the precision of the estimates. Future research with larger samples would be needed to draw more definitely conclusions about whether English language status modulates the Stroop interference effect.

## References
