---
title: "RMDL Assessment 2: Data Analysis Report"
authors:
  - name: B288421
bibliography: references.bib
execute: 
  echo: true
---

## Background

The Stroop task, first described by @stroop_studies_1935, is one of the most well known phenomenon in cognitive psychology. In a classic Stroop task setup, participants are presented with colored words printed in colored ink and asked to name the ink color while ignoring the word itself. The key finding is that participants are slower to respond when the word and ink color are incongruent compared to when they are congruent [@macleod_stroop_2006]. This difference in reaction time (RT) is known as Stroop interference, or the Stroop effect, and is thought to reflect the automaticity of word reading [@macleod_half_1991]. 

The Stroop effect has also been studied in bilingual speakers. @macleod_half_1991 reviews approximately two dozen studies on this topic. He notes that a dominant language has greater potential for interference than a nondominant one, and that the pattern of interference shifts as proficiency in the L2 increases. This suggests that the degree of Stroop interference in a given language reflects the extent to which word reading has become automatized in that language. Therefore, if word reading is less automatic in L2, the color word should interfere less with ink color identification, resulting in a smaller Stroop effect.

### Research Questions/Hypotheses

**Research questions**

-   **RQ 1:** Are participants slower when responding to incongruent trials compared to congruent trials?
-   **RQ 2:** Does the Stroop effect differ between L1 and L2 English speakers?

**Hypotheses**

-   **RH 1:** Participants will show slower reaction times on incongruent trials compared to congruent trials, reflecting the Stroop effect
-   **RH 2:** Since automatic word reading is stronger in a first language, L1 English speakers will show a larger Stroop effect compared to L2 English speakers.

## Methods

### Participants

25 participants took part in the study, and they were divided into L1 English speakers (n = 11) and L2 English speakers (n = 14) based on their self report. Participants were recruited through convenience sampling. 

### Materials

The experiment was designed and administered using Testable. The stimuli consisted of three color words (RED, BLUE, and GREEN) presented in colored ink (red, blue, or green). Each word appeared in either a congruent condition (word and ink color match) or an incongruent condition (word and ink color do not match). Each block contained 20 critical trials (10 congruent, 10 incongruent) and 12 filler trials consisting of non-color words (CAR, SHOE, HOUSE) presented in colored ink.

### Procedure

Participants completed the experiment online via Testable. They were instructed to respond to the ink color of the word by pressing the **LEFT** arrow key for red, the **UP** arrow key for blue, and the **RIGHT** arrow key for green as quickly and accurately as possible. Following the experimental trials, participants completed a short questionnaire asking whether English was their first language, their self-rated English proficiency, and whether they currently reside in an English-speaking country.

### Statistical analysis

Data were analyzed using Bayesian multilevel modeling in R using the `brms` package, and only critical trials were retained, and trials with RT below 200ms or above 2000ms were excluded as likely errors. A model was fitted with RT as the outcome variable using a log-normal family, with congruency, English language background, and their interaction as fixed effects, and varying intercepts and slopes for congruency by participant. The effect of congruency was used to assess whether a Stroop effect was present, and the interaction term was used to assess whether the Stroop effect differed between L1 and L2 speakers.

### Open research statement

The experimental trial file, raw data, and analysis code are available at \[<https://github.com/ww2526852/rdml-assignment2-repo>\]. The experiment was administered via Testable and can be accessed at \[<https://tstbl.co/661-427>\].

## Data

### Read and process the data

The data was downloaded from Testable in long format, with all 25 participants' trial results combined into a single CSV file. The file was read into R using read_csv().

Several processing steps were applied to prepare the data for analysis. First, each participant's response to the question "Is English your first language?" (row 82 in the trial file) was extracted and used to create a new column **english_status**, coded as **L1** for participant who answered "Yes" and **L2** for all other responses. This column identified the between-subjects grouping variable for the analysis.

The data was then filtered to retain only critical test trials (**condition1 == "critical"**), excluding practice and filler trials that are not relevant to the research question. Additionally, trials with RTs below 200ms or above 2000ms were excluded, as these are likely to reflect errors such as accidental key presses or inattentive responses rather than genuine task performance. The relevant columns in the cleaned dataset are as follows:

-   **filename:** unique identifier for each participant, based on their Testable result file
-   **condition2:** the congruency condition of each trial
-   **RT:** reaction time in milliseconds, measured from stimulus onset to key press
-   **correct:** accuracy of the responses, coded as 1 (correct) or 0 (incorrect)
-   **english_status:** English language background of the participant, coded as L1 (English is their first language) or L2 (English is not their first language)

```{r}
#| label: read-data

library(tidyverse)

#read raw data file
raw <- read_csv("data/661427_260428_102656_long.csv")

#extract rowNo 82 (L1 question) for each participant
l1_info <- raw |>
  filter(rowNo == 82) |>
  select(filename, l1_response = response)

#merge l1_info and create english_status column
raw <- raw |>
  left_join(l1_info, by = 'filename') |>
  mutate(
    english_status = ifelse(l1_response == 'Yes', 'L1', 'L2')
  )

#filter for critical trials and select relevant columns 
cleaned_data <- raw |>
  filter(type == 'test', condition1 == 'critical') |>
  select(filename, stim1, textColor, condition2, RT, correct, english_status) |>
  mutate(
    condition2 = factor(condition2, levels = c('congruent', 'incongruent')),
    english_status = factor(english_status, levels = c('L1', 'L2'))
  ) |>
  filter(RT > 200, RT < 2000)

```

{{< pagebreak >}}

### Plot the data

#### Plot 1: RT distribution

*Figure 1* shows the distribution of reaction time across congruency conditions and is faceted by *english_status.* The x-axis shows RT and the y-axis shows density. Congruency condition is indicated by fill color, with congruent trials shown in pink and incongruent trials shown in teal. The left panel shows L1 English speakers and the right panel shows L2 English speakers.

```{r}
library(ggplot2)

cleaned_data |>
  ggplot(aes(x = RT,fill = condition2)) + 
  geom_density(alpha=0.5) +
  facet_wrap(~english_status) +
  labs(
    x = 'RT',
    caption = 'Figure 1: Distribution of RT by congruency and English background',
    fill = 'Congruency'
  )
```

{{< pagebreak >}}

In both panels, the incongruent distribution is shifted rightward relative to the congruent distribution, indicating that participants were generally slower on incongruent trials regardless of language background, which is consistent with a Stroop effect. The shift appears visually larger in the L1 panel compared to the L2 panel, suggesting that the effect may be greater for L1 English speakers. 

#### Plot 2: Mean RT

*Figure 2* shows the mean reaction time for each congruency condition, with congruency on the x-axis and mean RT on the y-axis. English background is indicated by color, with L1 speakers shown in pink and L2 speakers in teal.

```{r}

cleaned_data |>
  group_by(english_status,condition2) |>
  summarize(
    mean_RT = mean(RT),
    sd_RT = sd(RT)
  ) |>
  ggplot(aes(x = condition2, y = mean_RT,color = english_status, 
             group = english_status)) +
  geom_point(size = 3) +
  geom_line()+
  geom_errorbar(aes(ymin = mean_RT - sd_RT,
                    ymax = mean_RT + sd_RT),
                width = 0.1) +
  labs(
    caption = 'Figrue 2: Mean RT by congruency and English status',
    x = 'Congruency',
    y = 'Mean RT',
    color = 'English background'
  )

```

Both groups show a clear increase in mean RT from congruent to incongruent trials, consistent with a Stroop effect in both L1 and L2 English speakers. The slope of the increase appears steeper for L1 speakers compared to L2 speakers, suggesting a larger effect for the L1 group. L2 speakers show higher overall RT in both conditions, suggesting they are generally slower to respond.

#### Plot 3: Accuracy

*Figure 3* shows the mean accuracy for each combination of congruency condition and English background. Congruency is shown on the x-axis and mean accuracy on the y-axis, ranging from 0 to 1. English background is indicated by fill color, with L1 speakers in pink and L2 speakers in teal. Bars for each group are displayed side by side within each congruency condition.

```{r}

cleaned_data |>
  group_by(english_status,condition2) |>
  summarize(mean_acc = mean(correct)) |>
  ggplot(aes(x = condition2,y = mean_acc,fill = english_status)) +
  geom_col(position = 'dodge') +
  labs(
    caption = 'Figure 3: Accuracy by congruency and English background',
    x = 'Congruency',
    y = 'Mean accuracy',
    fill = 'English background'
  ) +
  ylim(0,1)
```

Accuracy is uniformly high across all conditions, with all bars approaching 1.0. A small decrease in accuracy is visible from congruent to incongruent trials in both groups, consistent with the expected direction of a Stroop interference effect. The difference between L1 and L2 speakers within each congruency condition is minimal.

{{< pagebreak >}}

### Summarize the data

The table below reports mean RT, SD of RT, mean accuracy, and number of trials for each combination of congruency condition and English background groups.

```{r}
summary_stats <- cleaned_data |>
  group_by(english_status,condition2) |>
  summarize(
    mean_RT = mean(RT),
    sd_RT = sd(RT),
    accuracy = mean(correct),
    n_trials = n()
  )

summary_stats
```

L1 English speakers showed a mean RT of 607ms on congruent trials and 742ms on incongruent trials, reflecting a Stroop effect of approximately 135ms. L2 English speakers showed a mean RT of 677ms on congruent trials and 770ms on incongruent trials, reflecting a Stroop effect of approximately 93ms. Both groups show the expected pattern of slower responses on incongruent trials. The Stroop effect appears numerically larger for L1 English speakers than L2 English speakers. Accuracy was high across all conditions, exceeding 0.93 in all groups.

{{< pagebreak >}}

## Regression modelling

To investigate whether English background modulates the Stroop effect, a Bayesian multilevel model was fitted using the `brms` package in R with a log-normal family. Reaction times were used as the outcome variable. The model included congruency (congruent vs. incongruent), English background (L1 vs. L2), and their interaction as fixed effects.

The interaction term is of primary interest, as it tests whether the size of the Stroop effect differs between L1 and L2 English speakers. Varying intercepts and varying slopes for congruency were included by participant, allowing baseline RT and the congruency effect to differ across individuals.

```{r}
library(brms)

rt_bm <- brm(
  RT ~ condition2 * english_status + 
    (1 + condition2 | filename),
  family = lognormal,
  data = cleaned_data,
  cores = 4,
  seed = 1234,
  file = "data/cache/rt_bm"
)

summary(rt_bm)

```

{{< pagebreak >}}

The intercept represents the estimated mean RT for L1 English speakers on congruent trials (6.35, 95% CrI\[6.26, 6.45\]). The effect of congruency for L1 speakers was 0.18 units (95% CrI \[0.12, 0.25\]), indicating that incongruent trials were credibly slower than congruent trials. L2 speakers were estimated to be slightly slower overall than L1 speakers (0.11, 95% CrI\[-0.03, 0.24\]), though this difference was not conclusive as the CrI includes 0.

The interaction term between congruency and English background (-0.06, 95% CrI\[-0.15, 0.03\]) is of primary interest because it tests whether the Stroop effect differs between groups. The negative estimate indicates that L2 speakers show a smaller Stroop interference effect than L1 speakers, consistent with the predicted direction. However, the 95% CrI includes 0, meaning that there is no conclusive evidence that the two groups differ in the magnitude of the Stroop effect. The random effect indicates meaningful variability in baseline RT across participants (sd(Intercept) = 0.18), confirming that individuals differ in their overall response speed, while variability in the Stroop effect across participants was small (sd(condition2incongruent) = 0.05).

{{< pagebreak >}}

### Model plots

*Figure 4* shows the model-estimated mean RT for congruent and incongruent trials, with congruency on the x-axis and RT on the y-axis. English background is indicated by color, with L1 speakers in pink and L2 speakers in teal.

```{r}

plot(
  conditional_effects(rt_bm, 'condition2:english_status'),
  points = FALSE
) 
```

*Figure 4: Model-estimated RT by congruency and English background*

Both groups show an increase in model-estimated RT from congruent to incongruent trials, consistent with a Stroop interference effect in both L1 and L2 English speakers. The increase appears slightly larger for L1 speakers than L2 speakers, reflecting the numerically larger interference effect observed in the model coefficients.

*Figure 5* shows the posterior distributions of the Stroop effect for L1 speakers (green), L2 speakers (blue), and the difference between the two groups (pink). The x-axis show the estimated Stroop effect and the y-axis shows density. The vertical black line at 0 serves as a reference point: distributions entirely to the right of this line indicates a credibly positive effect, while distributions crossing the line indicate uncertainty about the direction of the effect.

```{r}
rt_draws <- as_draws_df(rt_bm, '^b', regex = TRUE)

rt_draws <- rt_draws |>
  mutate(
    stroop_L1 = b_condition2incongruent,
    stroop_L2 = b_condition2incongruent + `b_condition2incongruent:english_statusL2`,
    group_difference = stroop_L2 - stroop_L1
  )

rt_stroop_diff <- rt_draws |>
  select(stroop_L1, stroop_L2, group_difference) |>
  pivot_longer(everything())

rt_stroop_diff |>
  ggplot(aes(value, fill = name)) +
  geom_vline(xintercept = 0) +
  geom_density(alpha = 0.5)
```

*Figure 5: Posterior distributions of the Stroop effect for L1 and L2 speakers and their difference*

The posterior distributions for both **stroop_L1** and **stroop_L2** lie entirely to the right of 0, confirming that both groups show a credible Stroop interference effect. The **stroop_L1** distribution is shifted slightly further right than the **stroop_L2** distribution, and the **group_difference** distribution is centered to the left of 0, indicating that the Stroop effect is estimated to be smaller for L2 than L1 speakers. However, the distribution crosses 0, reflecting uncertainty about whether this difference is genuine.

```{r}

rt_stroop_diff |>
  group_by(name) |>
  summarise(
    q5  = round(quantile(value, 0.05), 3),
    q10 = round(quantile(value, 0.10), 3),
    q20 = round(quantile(value, 0.20), 3),
    q80 = round(quantile(value, 0.80), 3),
    q90 = round(quantile(value, 0.90), 3),
    q95 = round(quantile(value, 0.95), 3)
  )
```

The Stroop effect for L1 English speakers was credibly positive, with a 90% CrI of \[0.141, 0.226\] and a 95% CrI of \[0.127, 0.237\], both excluding 0. This provides clear evidence of a Stroop interference effect in L1 English speakers. Similarly, the Stroop effect for L2 English speakers was also credibly positive, with a 90% CrI of \[0.085, 0.159\] and 95% CrI of \[0.074, 0.170\], both excluding 0, indicating a Stroop interference effect in L2 English speakers as well.

The group difference was largely negative and consistent with the prediction that L1 English speakers will show greater interference than L2 English speakers. The 80% CrI of [-0.099, -0.025] and 90% CrI of [-0.117, -0.005] provide evidence of a group difference. However, the 95% CrI of [-0.134, 0.011] includes 0, meaning the evidence is not conclusive at the highest level of confidence examined.

{{< pagebreak >}}

## Discussion

The results provided clear evidence in support of RH 1. Both L1 and L2 English speakers showed credibly slower reaction times on incongruent trials compared to congruent trials. This is consistent with the well-established Stroop interference effect, where the automatic reading of a colored word interferes with the identification of its ink color. 

With respect RQ 2, the results provided weak to moderate evidence consistent with RH 2. The group difference was in the predicted direction, with L1 English speakers showing a larger Stroop effect than L2 English speakers. The 80% and 90% credible intervals for the group difference excluded 0, suggesting that at these levels of confidence, the Stroop effect is smaller for L2 and L1 speakers. However, the 95% credible interval included 0, and given the small sample size, these results should be interpreted carefully.

This inconclusiveness can be attributed to several factors. First, the sample size was small, which limited the precision of the estimates. Second, the majority of the L2 participants self-rated their English proficiency as advanced, suggesting a high degree of English language experience within the L2 group. This may have attenuated any group difference as highly proficient L2 speakers are likely to have developed more automatized word reading in English, which would have brought their Stroop effect closer in size to that of L1 speakers. 

## References
